A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-14 text/survival-methods.Rmd
full edit (dennis), add PCAWG pathway analysis reference to intro, update word template to fix spacing before and after subsection headers

In order to understand how the diagnosis of cancer changes the expectation for
longevity of patients, we first began by modeling the longevity of the overall
population of the US based on CDC life tables. Although childhood cancers exist
and are tragic, most cancer patients have diagnosis later in life. We modeled
baseline longevity using a risk rate function:

@@

Time is split into two components. $A$ going from 0 up to the age of the patient
at diagnosis and then remaining fixed, whereas $t$ starts at 0 upon diagnosis
and represents the days since diagnosis, a field recorded in the dataset. $D(t)$
is an indicator for whether diagnosis has occurred or not (before diagnosis D(t)
= 0, after it equals 1). The model essentially assumes that the risk of dying
per year at age 20 years is $R_{20}$ and that it increases exponentially with a
constant rate $a$ up to diagnosis, and then with a new rate $ka$
thereafter. Though the model does extend back to earlier ages, relatively few
patients were under age 20.

Given this risk per unit time, the probability to die at time $T+dT$ is the
probability to survive to time $T$ which is $(1-P(T))$ and then die in the
remaining interval which is $R(T)dT$

@@

The product $k_{tis}k_{cl}$ is obviously symmetric between the two factors. To
disambiguate the meaning of the two, the prior distribution of $k_{cl}$ has peak
probability at 1 and a relatively narrow width, due to the fact that it is a
tissue specific multiplicative modifying factor, this allows the overall
magnitude of k to be primarily determined by the $k_{tis}$ value whose prior is
significantly less constrained so that there is a wider range of dangers across
tissue types.

@@

Using this model, we split the data by Male/Female sex, and by White, vs Black
race as these were the main well-identified categories available to us in the
dataset. Sex is known to be a risk factor for death, with males dying at
slightly higher rates for all ages of interest in our dataset. Black vs White
race is an important category due to differences in access to care and other
socioeconomic and education related factors. However, due to the low number of
Black patient tumors, we use ethnicity only to identify the baseline risk
function, while the tissue and class specific multipliers are not ethnicity
specific.

>>>>

In order to understand how the diagnosis of cancer changes the expectation for
longevity of patients, we first modeled the longevity of the overall
population of the U.S. based on CDC life tables. Most TCGA cases are diagnosed late in life, so we fit a model that emphasizes accuracy in the right tail of the distribution. We modeled
baseline longevity using a risk rate function:

@@

Time is split into two components. $A$ going from 0 up to the age of the patient
at diagnosis and then remaining fixed, whereas $t$ starts at 0 upon diagnosis
and represents the days since diagnosis, a field recorded in the dataset. $D(t)$
is an indicator for whether diagnosis has occurred or not (before diagnosis $D(t) = 0$, 
after $D(t) = 1$). The model assumes that the risk of death
is $R_{20}$ at age 20 years and increases exponentially with a
constant rate $a$ up to the age at diagnosis. Risk of death assumes a new rate $ka$
thereafter. Though the model extends to earlier ages, relatively few
patients were under age 20.

Given this risk per unit time, the probability of death at time $T+dT$ is the
probability to survive to time $T$ which is $(1-P(T))$ and then die in the
remaining interval which is $R(T)dT$

@@

The product $k_{tis}k_{cl}$ is obviously symmetric between the two factors. To
disambiguate the meaning of the two, the prior distribution of $k_{cl}$ has peak
probability at 1 and a relatively narrow width, due to the fact that it is a
tissue specific multiplicative modifying factor. This allows the overall
magnitude of $k$ to be primarily determined by the $k_{tis}$ value whose prior is
significantly less constrained so that there is a wider range of risk across
tissue types.

@@

Using this model, we split the data by Male/Female sex, and by White, _vs._ Black
race as these were the main well-identified categories available to us in the
dataset. Sex is known to be a risk factor for death, with males dying at
slightly higher rates for all ages of interest in our dataset. Black _vs._ White
race is an important category due to differences in access to care and other
socioeconomic and education related factors. However, due to the low number of
Black patient tumors, we use ethnicity only to identify the baseline risk
function, while the tissue and class specific multipliers are not ethnicity
specific.
